A Gaussian process model of human electrocorticographic data

### 2020-02-13 paper/main.tex
backing up edits...

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original
sample rate) to 250 Hz.  (This downsampling step served to both
normalize for differences in sampling rates across patients and to
ease the computational burden of our subsequent analyses.)  We then
excluded any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited an maximum kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A); for Dataset 2, this yielded clean
recordings from 5023 electrodes from 78 patients.  Each individual
patient contributes electrodes from a limited set of brain locations,
which we localized in a common space~\citep[MNI152;][]{GrabEtal06}; an
example Dataset 1 patient's 54 electrodes that passed the above
kurtosis threshold test are highlighted in black and red.

The recording from a given electrode is maximally informative about
the activity of the neural tissue immediately surrounding its
recording surface.  However, brain regions that are distant from the
recording surface of the electrode also contribute to the recording,
albeit (ceteris paribus) to a much lesser extent.  One mechanism
underlying these contributions is volume conduction.  The precise rate
of falloff due to volume conduction (i.e., how much a small volume of
brain tissue at location $x$ contributes to the recording from an
electrode at location $\eta$) depends on the size of the recording
surface, the electrode's impedance, and the conductance profile of the
volume of brain between $x$ and $\eta$.  As an approximation of this
intuition, we place a Gaussian radial basis function (RBF) at the
location $\eta$ of each electrode's recording surface
(Fig.~\ref{fig:methods}B).  We use the values of the RBF at any brain
location $x$ as a rough estimate of how much structures around $x$
contributed to the recording from location $\eta$:
\begin{align}
  \mathrm{rbf}(x|\eta,\lambda) & =
  \mathrm{exp}\left\{ -\frac{||x - \eta||^2}{\lambda} \right\},\label{eqn:rbf}
\end{align}
where the width variable $\lambda$ is a parameter of the algorithm
(which may in principle be set according to location-specific tissue
conductance profiles) that governs the level of spatial smoothing.  In
choosing $\lambda$ for the analyses presented here, we sought to
maximize spatial resolution (which implies a small value of $\lambda$)
while also maximizing the algorithm's ability to generalize to any
location throughout the brain, including those without dense electrode
coverage (which implies a large value of $\lambda$).  Here we set
$\lambda = 20$, guided in part by our prior work~\citep{MannEtal14b,
  MannEtal18}, and in part by examining the brain coverage with
non-zero weights achieved by placing RBFs at each electrode location
in Dataset 1 and taking the sum (across all electrodes) at each voxel
in a 4~mm$^3$ MNI brain.  (We then held $\lambda$ fixed for our
analyses of Dataset 2.)  We note that this value could in theory be
further optimized, e.g., using cross validation or a formal
model~\citep[e.g.,][]{MannEtal18}.

@@

\begin{align}
  \bar{C}_{s} &=
  \mathrm{r}(\frac{1}{n}(\sum_{k=1}^{n}\mathrm{z}(C_{s,k}))),\label{eqn:inter_corr}~\mathrm{where}\\
\mathrm{z}(r) &= \frac{\log(1+r) - \log(1 - r)}{2}~\mathrm{is~the~Fisher}~z \mathrm{-transformation~and}\label{eqn:fishersz}\\
\mathrm{z}^{-1}(z) = \mathrm{r}(z) &= \frac{\exp(2z) - 1}{\exp(2z) + 1}\mathrm{~is~its~inverse}.\label{eqn:invfishersz}
\end{align}
Next, we use Equation~\ref{eqn:rbf} to construct a number of
to-be-estimated locations by number of patient electrode locations
weight matrix, $W_s$.  Specifically, $W_s$ approximates how informative the
recordings at each location in $R_s$ are in reconstructing activity at
each location in $\bar{R}$, where the contributions fall off with an RBF according to
the distances between the corresponding locations:

@@

Let $Y_{s,k,\alpha_s}$ be the number-of-timepoints ($T$) by
$\left|\alpha_s\right|$ matrix of (observed) voltages from the electrodes in
$\alpha_s$ during session $k$ from patient $s$. Then we can estimate the
voltage from patient $s$'s $k^{th}$ session at the locations in
$\beta_s$ using~\cite{Rasm06}:

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, we note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., an example patient in Dataset 1, $T =
12786750$). However, because Equation~\ref{eqn:reconstruction} is time
invariant, we may compute $Y_{s,k,\beta_s}$ in a piecewise
manner by filling in $Y_{s,k,\beta_s}$ one row at a time (using the
corresponding samples from $Y_{s, k, \alpha_s}$).

@@

As a stricter benchmark, we compared the quality of these
across-participant reconstructions (i.e., computed using a correlation
model learned from other patients' data) to reconstructions generated
using a correlation model trained using the in-patient's data.  In
other words, for this within-patient benchmark analysis we estimated
$\hat{C}_{s}$ (Eqn.~\ref{eqn:subj_corrmat}) for each patient in turn,
using recordings from all of that patient's electrodes except at the
location we were reconstructing.  These within-patient reconstructions
serve as an estimate of how well data from all of the other electrodes
from that single patient may be used to estimate held-out data from
the same patient.  This allows us to ask how much information about
the activity at a given electrode might be inferred through (a) volume
conductance or other sources of ``leakage'' from activity patterns
measured from the patient's other electrodes and (b) across-electrode
correlations learned from that single patient.  As shown in
Figure~\ref{fig:corrmap}A (gray histogram), the distribution of
within-patient correlations was centered well above zero (mean: 0.32;
$t$-test comparing mean of distribution of $z$-transformed average
patient correlation coefficients to 0: $t(66) = 15.16, p < 10^{-10}$).
However, the across-patient correlations were substantially higher
($t$-test comparing average $z$-transformed within versus across
patient electrode correlations: $t(66) = 9.17, p < 10^{-10}$).  This
is an especially conservative test, given that the across-patient
SuperEEG reconstructions exclude (from the correlation matrix
estimates) all data from the patient whose data is being
reconstructed.  We repeated each of these analyses on a second
independent dataset and found similar results
(Fig.~\ref{fig:corrmap}B, D; within versus across reconstruction
accuracy: $t(77) = 11.25, p < 10^{-10}$). We also replicated this result
separately for each of the two experiments from Dataset 2
(Fig.~\perexpcorrmaps).  This overall finding, that reconstructions of
held-out data using correlation models learned from \textit{other}
patient's data yield higher reconstruction accuracy than correlation
models learned from the patient whose data is being reconstructed, has
two important implications.  First, it implies that distant electrodes
provide additional predictive power to the data reconstructions beyond
the information contained solely in nearby electrodes.  (This follows
from the fact that each patient's grid, strip, and depth electrodes
are implanted in a unique set of locations, so for any given electrode
the closest electrodes in the full dataset tend to come from the
same patient.)  Second, it implies that the spatial correlations
learned using the SuperEEG algorithm are, to some extent, similar
across people.

@@

To test the task-specificity of the SuperEEG-derived correlation
models, we restricted the dataset to the 24 patients that participated
in both experiments and repeated the above within- and across-patient cross validation
procedures separately for Experiment 1 and Experiment 2 data from
Dataset 2.  We then compared the reconstruction accuracies for held-out
electrodes, for models trained within versus across the two
experiments, or combining across both experiments
(Fig.~\perexptaskrecon).  In every case we found that across-patient
models trained using data from all other patients out-performed
within-patient models trained on data only from the subject
contributing the given electrode ($t$s$(23) > 6.50, p$s$ < 10^{-5}$).  All
reconstruction accuracies also reliably exceeded chance performance
($t$s$(23) > 8.00, p$s$ < 10^{-8}$).  Average reconstruction accuracy was
highest for the across-patient models limited to data from the same
experiment (mean accuracy: $0.68$); next-highest for the
models that combined data across both experiments (mean
accuracy: $0.61$); and lowest for models trained across tasks (mean
accuracy: $0.47$).  This result also held for each of the Dataset 2
experiments individually (Fig.~\perexptaskreconseparated).  Taken
together, these results indicate that there are reliable commonalities
in the spatial correlations of full-brain activity across tasks, but
that there are also reliable differences in these spatial correlations
across tasks.  Whereas reconstruction accuracy benefits from
incorporating data from other patients, reconstruction accuracy is
highest when constrained to within-task data, or data that includes a
variety of tasks (e.g., Dataset 1, or combining across the two Dataset
2 experiments).

Although both datasets we examined provide good full-brain coverage
(when considering data from every patient; e.g.\
Fig.~\ref{fig:corrmap}C, D), electrodes are not placed uniformly
throughout the brain.  For example, electrodes are more likely to be
implanted in regions like the medial temporal lobe (MTL), and are
rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel
MNI152 brain, we computed the proportion of electrodes in the dataset
that were contained within a 20 MNI unit radius sphere centered on
that voxel.  We defined the \textit{density} at that location as this
proportion.  Across Datasets 1 and 2, the electrode placement
densities were similar (correlation by voxel:
$r = 0.6, p < 10^{-10}$).  We wondered whether regions with good
covererage might be associated with better reconstruction accuracy
(e.g. Fig.~\ref{fig:corrmap}C and D indicate that many electrodes in the
MTL have relatively high reconstruction accuracy, and occipital
electrodes tend to have relatively low reconstruction accuracy).  To
test whether this held more generally across the entire brain, for
each dataset we computed the electrode placement density for each
electrode from each patient (using the proportion of \textit{other}
patients' electrodes within 20 MNI units of the given electrode).  We
then correlated these density values with the across-patient
reconstruction accuracies for each electrode.  We found a negative
correlation between reconstruction accuracy and density for Dataset 1,
but was unreliable for Dataset 2 (Dataset 1: $r = -0.21, p = 0.05$; Dataset 2:
$r = -0.30, p = 0.15$).  This indicates that the reconstruction
accuracies we observed are not driven by sampling density, but
rather may also reflect higher order properties of neural dynamics
such as functional correlations between distant
voxels~\citep{BetzEtal17b}.

>>>>

We first applied fourth order Butterworth notch filter to remove 60 Hz
($\pm$ .5 Hz) line noise from every recording (from every electrode).
Next, we downsampled the recordings (regardless of the original sample
rate) to 250 Hz.  (This downsampling step served to both normalize for
differences in sampling rates across patients and to ease the
computational burden of our subsequent analyses.)  We then excluded
any electrodes that showed putative epileptiform activity.
Specifically, we excluded from further analysis any electrode that
exhibited a maximum kurtosis of 10 or greater across all of that
patient's recording sessions.  We also excluded any patients with
fewer than 2 electrodes that passed this criteria, as the SuperEEG
algorithm requires measuring correlations between 2 or more electrodes
from each patient.  For Dataset 1, this yielded clean recordings from
4168 electrodes implanted throughout the brains of 67 patients
(Fig.~\ref{fig:methods}A, colored dots); for Dataset 2, this yielded
clean recordings from 5023 electrodes implanted in 78 patients.  Each
individual patient contributed electrodes from a limited set of brain
locations, which we localized in a common
space~\citep[MNI152;][]{GrabEtal06}; an example Dataset 1 patient's 54
electrodes that survived the kurtosis thresholding procedure are
highlighted in black and red (Fig.~\ref{fig:methods}A).

The recording from a given electrode is maximally informative about
the activity of the neural tissue immediately surrounding its
recording surface.  However, brain regions that are distant from the
recording surface of the electrode also contribute to the recording,
albeit (ceteris paribus) to a much lesser extent.  One mechanism
underlying these contributions is volume conduction.  The precise rate
of falloff due to volume conduction (i.e., how much a small volume of
brain tissue at location $x$ contributes to the recording from an
electrode at location $\eta$) depends on the size of the recording
surface, the electrode's impedance, and the conductance profile of the
volume of brain between $x$ and $\eta$.  As an approximation of this
intuition, we place a Gaussian radial basis function (RBF) at the
location $\eta$ of each electrode's recording surface
(Fig.~\ref{fig:methods}B).  We use the values of the RBF at any brain
location $x$ as a rough estimate of how much structures around $x$
contributed to the recording from location $\eta$:
\begin{align}
  \mathrm{rbf}(x|\eta,\lambda) & =
  \mathrm{exp}\left\{ -\frac{||x - \eta||^2}{\lambda} \right\},\label{eqn:rbf}
\end{align}
where the width variable $\lambda$ is a parameter of the algorithm
(which may in principle be set according to location-specific tissue
conductance profiles) that governs the level of spatial smoothing.  In
choosing $\lambda$ for the analyses presented here, we sought to
maximize spatial resolution (which implies a small value of $\lambda$)
while also maximizing the algorithm's ability to generalize to any
location throughout the brain, including those without dense electrode
coverage (which implies a large value of $\lambda$).  Here we set
$\lambda = 20$, guided in part by our prior related work~\citep{MannEtal14b,
  MannEtal18}, and in part by examining the brain coverage with
non-zero weights achieved by placing RBFs at each electrode location
in Dataset 1 and taking the sum (across all electrodes) at each voxel
in a 4~mm$^3$ MNI brain.  (We then held $\lambda$ fixed for our
analyses of Dataset 2.)  We note that this value could in theory be
further optimized, e.g., using cross validation or a formal
model~\citep[e.g.,][]{MannEtal18}.

@@

\begin{align}
  \bar{C}_{s} &=
  \mathrm{r}\left(\frac{1}{n}\left(\sum_{k=1}^{n}\mathrm{z}(C_{s,k})\right)\right),\label{eqn:inter_corr}~\mathrm{where}\\
\mathrm{z}(r) &= \frac{\log(1+r) - \log(1 - r)}{2}~\mathrm{is~the~Fisher}~z \mathrm{-transformation~and}\label{eqn:fishersz}\\
\mathrm{z}^{-1}(z) = \mathrm{r}(z) &= \frac{\exp(2z) - 1}{\exp(2z) + 1}\mathrm{~is~its~inverse}.\label{eqn:invfishersz}
\end{align}
Next, we use Equation~\ref{eqn:rbf} to construct a number of
to-be-estimated locations by number of patient electrode locations
weight matrix, $W_s$.  Specifically, $W_s$ approximates how informative the
recordings at each location in $R_s$ are in reconstructing activity at
each location in $\bar{R}$, where the contributions fall off with an RBF according to
the distances between the corresponding locations:

@@

Let $Y_{s,k,\alpha_s}$ be the number-of-timepoints ($T$) by
$\left|\alpha_s\right|$ matrix of (observed) voltages from the electrodes in
$\alpha_s$ during session $k$ from patient $s$. Then we can estimate the
voltage from patient $s$'s $k^{th}$ session at the locations in
$\beta_s$ as follows~\citep{Rasm06}:

@@

We designed our approach to be agnostic to electrode impedances, as
electrodes that do not exist do not have impedances.  Therefore our
algorithm recovers voltages in standard deviation ($z$-scored) units
rather than attempting to recover absolute voltages. (This property
reflects the fact that $\hat{K}_{\beta_s, \alpha_s}$ and
$\hat{K}_{\alpha_s, \alpha_s}$ are correlation matrices rather than
covariance matrices.)  Also, we note that
Equation~\ref{eqn:reconstruction} requires computing a $T$ by $T$
matrix, which can become computationally intractable when $T$ is very
large (e.g., an example patient in Dataset 1, $T = 12,786,750$; also
see Fig.~\suppstats, Panels E and K). However, because
Equation~\ref{eqn:reconstruction} is time invariant, we may compute
$Y_{s,k,\beta_s}$ in a piecewise manner by filling in
$Y_{s,k,\beta_s}$ one row at a time (using the corresponding samples
from $Y_{s, k, \alpha_s}$).

@@

Next, we compared the quality of these
across-participant reconstructions (i.e., computed using a correlation
model learned from other patients' data) to reconstructions generated
using a correlation model trained using the in-patient's data.  In
other words, for this within-patient benchmark analysis we estimated
$\hat{C}_{s}$ (Eqn.~\ref{eqn:subj_corrmat}) for each patient in turn,
using recordings from all of that patient's electrodes except at the
location we were reconstructing.  These within-patient reconstructions
serve as an estimate of how well data from all of the other electrodes
from that single patient may be used to estimate held-out data from
the same patient.  This allows us to ask how much information about
the activity at a given electrode might be inferred through (a) volume
conductance or other sources of ``leakage'' from activity patterns
measured from the patient's other electrodes and (b) across-electrode
correlations learned from that single patient.  As shown in
Figure~\ref{fig:corrmap}A (gray histogram), the distribution of
within-patient correlations was centered well above zero (mean: 0.32;
$t$-test comparing mean of distribution of $z$-transformed average
patient correlation coefficients to 0: $t(66) = 15.16, p < 10^{-10}$).
However, the across-patient correlations were substantially higher
($t$-test comparing average $z$-transformed within versus across
patient electrode correlations: $t(66) = 9.17, p < 10^{-10}$).  This
is an especially conservative test, given that the across-patient
SuperEEG reconstructions exclude (from the correlation matrix
estimates) all data from the patient whose data is being
reconstructed.  We repeated each of these analyses on a second
independent dataset and found similar results
(Fig.~\ref{fig:corrmap}B, D; within versus across reconstruction
accuracy: $t(77) = 11.25, p < 10^{-10}$). We also replicated this result
separately for each of the two experiments from Dataset 2
(Fig.~\perexpcorrmaps).  This overall finding, that reconstructions of
held-out data using correlation models learned from \textit{other}
patient's data yield higher reconstruction accuracy than correlation
models learned from the patient whose data is being reconstructed, has
two important implications.  First, it implies that distant electrodes
provide additional predictive power to the data reconstructions beyond
the information contained solely in nearby electrodes.  (This follows
from the fact that each patient's grid, strip, and depth electrodes
are implanted in a unique set of locations, so for any given electrode
the closest electrodes in the full dataset tend to come from the
same patient.)  Second, it implies that the spatial correlations
learned using the SuperEEG algorithm are, to some extent, similar
across people.

@@

To test the task-specificity of the SuperEEG-derived correlation
models, we restricted the dataset to the 24 patients that participated
in both experiments and repeated the above within- and across-patient
cross validation procedures separately for Experiment 1 and Experiment
2 data from Dataset 2.  We then compared the reconstruction accuracies
for held-out electrodes, for models trained within versus across the
two experiments, or combining across both experiments
(Fig.~\perexptaskrecon).  In every case we found that across-patient
models trained using data from all other patients out-performed
within-patient models trained on data only from the subject
contributing the given electrode ($t$s$(23) > 6.50, p$s$ < 10^{-5}$).
All reconstruction accuracies also reliably exceeded chance
performance ($t$s$(23) > 8.00, p$s$ < 10^{-8}$).  Average
reconstruction accuracy was highest for the across-patient models
limited to data from the same experiment (mean accuracy: $0.68$);
next-highest for the models that combined data across both experiments
(mean accuracy: $0.61$); and lowest for models trained across tasks
(mean accuracy: $0.47$).  This pattern of results also held for each
of the Dataset 2 experiments individually
(Fig.~\perexptaskreconseparated).  Taken together, these results
indicate that there are reliable commonalities in the spatial
correlations of full-brain activity across tasks, but that there are
also reliable differences in these spatial correlations across tasks.
Whereas reconstruction accuracy benefits from incorporating data from
other patients, reconstruction accuracy is highest when constrained to
within-task data, or data that includes a variety of tasks (e.g.,
Dataset 1, or combining across the two Dataset 2 experiments).

Although both datasets we examined provide good full-brain coverage
(when considering data from every patient; e.g.\
Fig.~\ref{fig:corrmap}C, D), electrodes were not sampled uniformly
throughout the brain.  For example, electrodes are more likely to be
implanted in regions like the medial temporal lobe (MTL), and are
rarely implanted in occipital cortex (Fig.~\ref{fig:density}A, B).
Separately for each dataset, for each voxel in the 4~mm$^3$ voxel
MNI152 brain, we computed the proportion of electrodes in the dataset
that were contained within a 20 MNI unit radius sphere centered on
that voxel.  We defined the \textit{density} at that location as this
proportion.  Across Datasets 1 and 2, the electrode placement
densities were similar (correlation by voxel:
$r = 0.6, p < 10^{-10}$).  We wondered whether regions with good
covererage might be associated with better reconstruction accuracy.
For example, Figures~\ref{fig:corrmap}C and D indicate that many
electrodes in the MTL (which tends to be relatively densely sampled)
have relatively high reconstruction accuracy, and occipital electrodes
(which tends to be relatively sparsely sampled) tend to have
relatively low reconstruction accuracy.  To test whether this held
more generally across the entire brain, for each dataset we computed
the electrode placement density for each electrode from each patient
(using the proportion of \textit{other} patients' electrodes within 20
MNI units of the given electrode).  We then correlated these density
values with the across-patient reconstruction accuracies for each
electrode.  We found a statistically reliable negative correlation
between reconstruction accuracy and density for Dataset 1, but no
reliable correlation for Dataset 2 (Dataset 1: $r = -0.21, p = 0.05$;
Dataset 2: $r = -0.30, p = 0.15$).  This indicates that the
reconstruction accuracies we observed are \textit{not} driven solely
by sampling density, but rather may also reflect higher order
properties of neural dynamics such as functional correlations between
distant voxels~\citep{BetzEtal17b}.

%%%%%%%%%%%%%%%%%%%%% MOVE FREQEUNCY STUFF HERE
